Convergence Studies and Effective Alphas for Disks with Gravitational Instabilities. \\ I. The Constant Cooling Case

### 2010-12-14 Paper_1.tex
Edited through end of section 2.
	modified:   Paper_1.tex
	modified:   general.bib

\documentclass[manuscript]{aastex} % one-column, double-spaced document
%\documentclass[preprint2]{aastex}  %% double-column, single-spaced document
% \documentclass[12pt,preprint]{aastex}
% \documentclass[aaspp4]{aastex}
% \documentclass[aas2pp4]{aastex}
% \documentclass[aj_pt4]{aastex}
% \documentclass[apjpt4]{aastex}
% \documentstyle[12pt,aasms4]{article}
% \documentclass[12pt]{article}
%\documentstyle[12pt]{article}
% \tighten
%
%
\newcommand{\ie}{{\it i.e.}}
\newcommand{\cf}{{\it cf.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}
\newcommand{\rd}{{\rm d}}
\newcommand{\ag}{\mbox{ \raisebox{-.8ex}{$\stackrel{\textstyle >}{\sim}$} }}
\newcommand{\al}{\mbox{ \raisebox{-.8ex}{$\stackrel{\textstyle <}{\sim}$} }}
\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{tomsc@astro.indiana.edu}
\newcommand{\mso}{$M_{\odot}$ }
\newcommand{\rso}{$R_{\odot}$ }
\newcommand{\msoyr}{$M_{\odot}/{\rm year}$}

@@

We conducted a 3D hydrodynamic convergence study of a protostellar disk
subject to a constant global cooling time 
and to gravitational instabilities (GIs). 
The disk was followed during its asymptotic state, where
heating and cooling are in balance.
Four simulations were conducted, identical except for the number of azimuthal grid 
points, $l_{max} = $ 64, 128, 256 and 512, used in the calculations. 
The strengths of the azimuthal modes, gravitational torques, and the effective \citet{shakura1973} $\alpha$ 
arising from gravitational stresses were determined for each resolution.  These
were used to examine the resolution-dependence of each quantity
and to evaluate both the converged modal power distribution and the effective value of $\alpha$
as $l_{max} \rightarrow \infty$.   Power was found in all modes and 
the power summed over all modes was essentially independent of resolution.
The modal distribution of power, however, was significantly affected by resolution.  
The size of the azimuthal grid, $l_{max}$,
limits the number of modes to $l_{max}/2$.  Higher resolution simulations therefore
have more modes available to them.  Power is spread across all available modes.  As a consequence,
higher resolution simulations have a larger fraction of their total power in high-order modes.
High-order modes
manifest themselves more locally than low-order modes, therefore 
greater resolution leads to smaller gravitational stresses.
As a result, higher resolution simulations experience 
weaker gravitational torques than lower resolution simulations.   The effective  $\alpha$
also depends upon the magnitude of the stresses.  Thus
$\alpha_{eff}$ is also a function of grid resolution, decreasing as  $l_{max}$ increases.
The converged $\alpha_{eff}$ is consistent with
predictions from an analytic local theory for thin disks by \citet{gammie2001}, 
but only over many dynamic times when averaged over a substantial volume of the disk.
Results obtained in the $l_{max} = 512$ simulations differ with converged values
by less than 20\%.

@@

%\keywords{ }

%%%%%% INTRODUCTION %%%%%%%%%%%%%%%
\section{Introduction}

Detailed three-dimensional hydrodynamic modeling of protostellar disks can provide considerable insight into the physical and thermal states of these disks.  However,  computing resources limit these studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime. The level of spatial resolution (and hence temporal resolution, due to the Courant condition) used in calculations can play a deterministic role in interpretations of models; while lower resolutions allow more extensive spatial and temporal studies, this benefit can come at the sacrifice of important physics.  Convergence studies provide a means 
of deriving important outcomes without the expense of very high resolution simulations.

@@

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep{nelson1998, pickett1998, lodato2004,boley2006,cossins2009}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, referred to as the {\it asymptotic state},
then power will exist at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In the asymptotic state, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.  Here convergence studies are important because 
higher order modes dissipate energy on shorter wavelengths than lower order modes.  The mix of low and high order modes available to a disk is set by the resolution of the simulation.

@@

Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance, i.e., disks subject to gravitoturbulence,
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, 
an $\alpha$-viscosity may be consistent
with transport in GI acitve disks for disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}.

\citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  
Assuming a razor thin gravitationally unstable disk where the heating due to local turbulent dissipation caused by
GIs was balanced by a local cooling rate of $t_{cool}$.
Gammie found
$\alpha_{eff} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1}$,
where $\gamma$ is the two dimensional adiabatic index and $t_{cool}$ is a cooling parameter. 
The two dimensional index relates
to the three dimensional index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived essentially the same result for their 3D study.
Motivated by
the \cite{laughlin1996}, \cite{gammie2001}, and \cite{lodato2004, lodato2005} results 
that demonstrated differences in how low and high order modes manifest themselves, Kratter et al. (2008) 
proposed $\alpha_{eff} = (\alpha_{short}^2 + \alpha_{long}^2)^{1/2}$, where $\alpha_{short}$ and $\alpha_{long}$
are functions of $Q$ and $M_d/M_*$ representing 
the different wavelength regimes expected to dominate for different values of the mass ratio and cooling rate. 
{\bf (Scott: need to add kratter2008 to the bib file.  The reference is Kratter, K.M., Matzner, C. D., Krumholz, M.R., 
2008. ApJ 681, 375)}
Fully 3D simulations \citep[e.g.][]{lodato2004, cossins2009} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{eff}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

We address these, and similar questions, using a grid-based, finite-difference 3D hydrodynamics code
with self gravity.
For computational grids of a fixed {\it physical} size,
grids with more elements provide finer meshes, and thus a more accurate solution. 
However, increased accuracy is offset by temporal sacrifices.
The number of equations solved at each timestep 
increases as the grid size is increased.  At the same time, the Courant condition 
forces shorter timesteps.  A mesh convergence study 
provides a common approach to balancing accuracy and computing resources.
In the following sections we follow the evolution of a disk subject to GIs
using four different
resolutions in the azimuthal direction.  The number of azimuthal grid points used in simulations
may be particularly important, because higher azimuthal resolution 
allows GI power to spread to higher-order modes 
that behave more locally. We test this by examining the power in various 
modes to see how they are affected by the choice of grid.
In the process, 
we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium; if so, to what level do these disks approximate the gravitoturbulence of \citet{gammie2001} and with what distribution of $\alpha_{eff}$; and to characterize how the GI cascades from low-order to high-order modes. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

@@

The initially unstable disk model passed through several phases of evolution \citep[see also][]{pickett2003, mejia2005}.  The initial disk is unstable to the growth of axisymmetric structures. During this axisymmetric phase, 
which lasts several ORPs, the disk cools and contracts, the contraction is small radially but dramatic in the vertical direction. Around 3-4 ORPs the instabilities begin grow to non-linear amplitudes. When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of a few ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the nonaxisymmetry in the disk.  Finally the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,64,128), (512,64,256), and (512,64,512). Because of the large computational overhead associated with starting each simulation from $t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase and it introduces a perturbation to the disk which takes several ORPs to subside. By 12 ORPs, the perturbation has dissipated and all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure 
%\ref{qcompareres}
\ref{fig:Final_Q} 
shows $Q$, as a function of radius, at $t \sim$ 18 ORP.  Table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to the growth of nonaxisymmetric structures over this full range.  Interior to $\sim$ 27 AU, the value of $Q$ shows no dependence on
the azimuthal resolution.  From 27 -- 50 AU, lower resolution simulations display lower values of $Q$.

@@

% \begin{deluxetable}{cccc} 
% \tablecolumns{4} 
% \tablewidth{0pc} 
% \tablecaption{Effective Alphas} 
% \tablehead{ 
% \colhead{}    &  \multicolumn{3}{c}{Non-shell Stars} &   \colhead{}   & 
%\multicolumn{3}{c}{Shell Stars} \\ 
%\cline{2-4} \cline{6-8} \\ 
%\cline{2-8} \\ 
% \colhead{Azimuthal Zones} & \colhead{$\alpha$}   & \colhead{$A_{tot}$}    
% & \colhead{$A_{2-5} / A_{tot}$} \\ 
%\colhead{} & \colhead{} &\colhead{} &  \colhead{angle} & \colhead{$10^{-28}$ ergs /}  \\
%\colhead{} & \colhead{} &\colhead{} &  \colhead{angle} 
%&  \multicolumn{2}{c}{$(10^{-28}$ ergs / cm$^3$ s sr)}  \\
% } 
% \startdata
% 64	 &    0.0587	 &        2.02 	 & 0.752	      \\
% 128	   &    0.0402	  &       1.99  	 & 0.608      \\
% 256	  &     0.0237	 &        1.87	 & 0.535     \\
% 512	  &     0.0241	 &        2.28	 & 0.467     \\
% \enddata 
% \tablecomments{
% Table 1. The $\alpha$ of figure 1, averaged over the radial region from 10 -- 40 AU, tabulated for different 
% azimuthal resolutions, along with amplitudes of Fourier components for non-axisymmetric density structures. 
% The radially averaged $\alpha$ predicted by Gammie (2001) are tablulated for comparison.
% }
% \end{deluxetable}

@@

\begin{figure}\label{Final_Q}
\plotone{figures/Figure1.eps}
\caption{Toomre Q distribution versus radius for each of the four resolutions at t $\sim$ 18 ORPs}
\end{figure}

\begin{figure}\label{fig:DensityPlots}
%\plotone{figures/Figure2.eps}
\caption
{
Midplane disk densities at $t \sim 18$ ORPs for simulations with 64, 128, 256 and 512 azimuthal grid elements.
The simulations differ only in the number of  azimuthal zones used in the calculations.  Densities out of the plane, along an azimuthal cut through the disk, are displayed below the face-on views.  The color scale is  logarithmic, and axis units are AU. 
}
\end{figure}
\newpage

\begin{figure}\label{fig:Am_vs_log_m}
%\plotone{figures/Figure3.eps}
\caption
{
The logarithm of the amplitudes for the global Fourier components of the non-axisymmetric density structure plotted for different $m$-values. Notice that the components with the highest global amplitudes are those with low $m$-values.  
}
\end{figure}

\begin{figure}
%\plotone{figures/Figure4.eps}
\caption
{
Convergence of fractional power.  The $x$-axis represents
the azimuthal ``spacing'' between grid points,  normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.
The $y$-axis displays the fraction of the total power in the low-order ($m = $ 2--7) Fourier modes. 
Also plotted is the Richardson Extrapolation for the asymptotic value  
$\langle A_{2-7} \rangle / \langle A_{tot} \rangle = 0.43$ as the grid spacing goes to zero.  
}
\end{figure}

\begin{figure}
%\plotone{figures/Figure5.eps}
\caption
{
Torque Array}
\end{figure}

%\begin{figure}
%\plotone{figures/Mode_CascadeRatios.pdf}
%\caption
%{
%Mode Cascade Ratios .  Note that by the end of the time frame shown, there is rough parity between 
%torques arising from tower order ($m = 2$ -- 5) mades and torques from higher order modes.}
%\end{figure}

\begin{figure}
\label{fig:alpha_v_radius}
%\plotone{figures/Figure6.eps}
\caption
{Effective $\alpha$'s for the $t_{cool} =$ 2 ORP, $\gamma = 5/3$, $\Sigma \sim r^{ -1}$ disk, averaged between 12 and 18 ORPs, for four different different azimuthal grid resolutions, given by the number of azimuthal grid-points Lmax. The parallel dashed lines show the predictions of Gammie (2001) for $\alpha$ (upper: strongly self-gravitating limit, lower: non-self-gravitating limit), while the dashed line at $\log(\alpha) = -1.22$ represents the maximum stress a disk can maintain without fragmenting (Rice et al. 2005) . 
}
\end{figure}
\newpage

\begin{figure}
%\plotone{figures/Figure7.eps}
\caption
{Alpha variations.}
\end{figure}

>>>>

\documentclass[manuscript]{aastex}

\newcommand{\ie}{{\it i.e.}}
\newcommand{\cf}{{\it cf.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}
\newcommand{\rd}{{\rm d}}
\newcommand{\ag}{\mbox{ \raisebox{-.8ex}{$\stackrel{\textstyle >}{\sim}$} }}
\newcommand{\al}{\mbox{ \raisebox{-.8ex}{$\stackrel{\textstyle <}{\sim}$} }}
\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{tomsc@astro.indiana.edu}
\newcommand{\mso}{$M_{\odot}$ }
\newcommand{\rso}{$R_{\odot}$ }
\newcommand{\msoyr}{$M_{\odot}/{\rm year}$}

@@

We conducted a convergence study of the azimuthal resolution in 3D hydrodynamic simulations of a protostellar disk susceptible to gravitational instabilities (GIs). 
The disk was primarily studied during its asymptotic state, when
heating and cooling are in balance. The cooling was treated as a constant global cooling time.
Four simulations were conducted, identical except for the number of azimuthal grid 
points, $l_{max} = $ 64, 128, 256 and 512, used in the calculations. 
The strengths of the azimuthal modes, gravitational torques, and the effective \citet{shakura1973} $\alpha$ 
arising from gravitational stresses were determined for each resolution.  We examined the resolution-dependence of each quantity
and evaluated both the converged modal amplitude distribution and the effective value of $\alpha$
as $l_{max} \rightarrow \infty$.   Non-axisymmetric amplitude was found in all modes and 
the amplitude summed over all modes was essentially independent of resolution.
The modal distribution of amplitudes, however, was significantly affected by resolution.  
The size of the azimuthal grid, $l_{max}$,
limits the number of modes that can be accurately measured to $l_{max}/2$.  Higher resolution simulations therefore
have more modes available to them.  Non-axisymmetric structure is present in all available modes.  As a consequence,
higher resolution simulations have a larger fraction of their total amplitude in high-order modes.
High-order modes
are more confined to act on local structure than low-order modes, therefore as the resolution increases and low-order amplitudes decrease the total gravitational stresses decreases as well.
As a result, higher resolution simulations experience 
weaker gravitational torques than lower resolution simulations.   The effective  $\alpha$
also depends upon the magnitude of the stresses.  Thus
$\alpha_{eff}$ is also a function of grid resolution, decreasing as  $l_{max}$ increases.
The converged $\alpha_{eff}$ is consistent with
predictions from an analytic local theory for thin disks by \citet{gammie2001}, 
but only over many dynamic times when averaged over a substantial volume of the disk.
Results obtained in the $l_{max} = 512$ simulations differ with converged values
by less than 20\%.

@@

\section{Introduction}

Detailed three-dimensional hydrodynamic modeling of protostellar disks has provided considerable insight into the physical and thermal states of these disks.  However,  computing resources limit these studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime. The level of spatial resolution (and hence temporal resolution, due to the Courant condition) used in calculations can play a deterministic role in interpretations of models; while lower resolutions allow more extensive spatial and temporal studies, this benefit can come at the sacrifice of important physics.  Convergence studies provide a means 
of deriving important outcomes without the expense of very high resolution simulations.

@@

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep{nelson1998, pickett1998, lodato2004,boley2006,cossins2009}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, referred to as the {\it asymptotic state},
then power will exist at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In the asymptotic state, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.  Here convergence studies are important because 
higher order modes dissipate energy on shorter wavelengths than lower order modes.  The mix of low and high order modes available to a disk is set by the resolution of the simulation. {\bf Multiple issues with this paragraph. Didn't want to rewrite unilaterally. 1) In the first couple sentences the description of mode growth needs some tweaking, many modes appear in the linear regime, and although we have seen amplitude at all m values I don't know that it's a given. 2) I wouldn't describe the structures formed by GI activity as ``strongly nonlinear turbulence''. 3) In reviewing Gammie 2001 it seems that he is actually referring to a turbulent cascade (a fact I hadn't fully appreciated until now) I think it's pretty clear that the effective alpha we see is, in fact, not due to a turbulent cascade. i.e. alpha is driven by low order coherent modes. We should discuss this at some point}

@@

Several studies support the idea that angular momentum transport by GIs in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance, i.e., disks subject to gravitoturbulence,
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, 
an $\alpha$-viscosity may be consistent
with transport in GI acitve disks for disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}.

\citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating caused by GIs was balanced by the local cooling rate.  
Assuming a razor thin gravitationally unstable disk where the heating caused by
GIs was balanced by a local cooling rate of $t_{cool}\Omega$.
Gammie found
$\alpha_{eff} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1}$,
where $\gamma$ is the two dimensional adiabatic index and $t_{cool}$ is a cooling parameter. 
The two dimensional index relates
to the three dimensional index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived a similar result when considering three dimensional protoplanetary disks.
Motivated by
the results from \cite{laughlin1996}, \cite{gammie2001}, and \cite{lodato2004, lodato2005}, which demonstrated differences in the manifestation of high and low-order, \citet{kratter2008} 
proposed $\alpha_{eff} = (\alpha_{short}^2 + \alpha_{long}^2)^{1/2}$. Here $\alpha_{short}$ and $\alpha_{long}$
are functions of $Q$ and $M_d/\left( M_d+M_*\right)$ representing 
the different wavelength regimes expected to dominate for different values of the mass ratio and cooling rate. 
Fully 3D simulations \citep[e.g.][]{lodato2004, cossins2009} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{eff}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

We address these, and similar questions, using a grid-based, finite-difference 3D hydrodynamics code
with self gravity.
For computational grids of a fixed {\it physical} size,
grids with more elements provide finer meshes, and thus a more accurate solution. 
However, increased accuracy is offset by computational penalties.
The number of equations solved at each timestep 
increases as the grid size is increased.  At the same time, the Courant condition 
forces shorter timesteps.  A mesh convergence study 
provides a common approach to balancing accuracy and computing resources.
In the following sections we follow the evolution of a disk subject to GIs
using four different
resolutions in the azimuthal direction.  The number of azimuthal grid points used in simulations
may be particularly important, because higher azimuthal resolution 
allows non-axisymmetric amplitude to spread to higher-order modes 
that behave more locally. We test this by examining the amplitude in various 
modes to see how they are affected by the choice of grid.
In the process, 
we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium If this is the case, we investigate how closely these disks approximate the gravitoturbulence of \citet{gammie2001} and examine the distribution of $\alpha_{eff}$. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

@@

The initially unstable disk model passed through several phases of evolution \citep[see also][]{pickett2003, mejia2005}.  The initial disk is unstable to the growth of non-axisymmetric structures. During the initial axisymmetric phase, 
which lasts several ORPs, the disk cools and contracts, the contraction is small radially but dramatic in the vertical direction. Around 3-4 ORPs the instabilities begin grow to non-linear amplitudes. When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of a few ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the nonaxisymmetry in the disk.  Finally, following this transition phase, the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,64,128), (512,64,256), and (512,64,512). Because of the computational overhead associated with running each simulation starting from $t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase. By 12 ORPs, all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{fig:Final_Q} shows $Q$, as a function of radius, at $t \sim$ 18 ORP.  Table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to the growth of nonaxisymmetric structures over this full range.  Interior to $\sim$ 27 AU, the value of $Q$ shows no dependence on
the azimuthal resolution.  From 27 -- 50 AU, lower resolution simulations display lower values of $Q$.

@@

@@

\begin{figure}
\plotone{figures/Figure1.eps}
\caption{Toomre Q distribution versus radius for each of the four resolutions at t $\sim$ 18 ORPs}
\label{fig:Final_Q}
\end{figure}

\begin{figure}
\plotone{figures/Figure2.eps}
\caption
{
Midplane disk densities at $t \sim 18$ ORPs for simulations with 64, 128, 256 and 512 azimuthal grid elements.
The simulations differ only in the number of  azimuthal zones used in the calculations.  Densities out of the plane, along an azimuthal cut through the disk, are displayed below the face-on views.  The color scale is  logarithmic, and axis units are AU. 
}
\label{fig:DensityPlots}
\end{figure}
\newpage

\begin{figure}
\plotone{figures/Figure3.eps}
\caption
{
The logarithm of the amplitudes for the global Fourier components of the non-axisymmetric density structure plotted for different $m$-values. Notice that the components with the highest global amplitudes are those with low $m$-values.  
}
\label{fig:Am_vs_log_m}
\end{figure}

\begin{figure}
\plotone{figures/Figure4.eps}
\caption
{
Convergence of fractional power.  The $x$-axis represents
the azimuthal ``spacing'' between grid points,  normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.
The $y$-axis displays the fraction of the total power in the low-order ($m = $ 2--7) Fourier modes. 
Also plotted is the Richardson Extrapolation for the asymptotic value  
$\langle A_{2-7} \rangle / \langle A_{tot} \rangle = 0.43$ as the grid spacing goes to zero.  
}
\end{figure}

\begin{figure}
\plotone{figures/Figure5.eps}
\caption
{
Torque Array}
\end{figure}

\begin{figure}
\label{fig:alpha_v_radius}
\plotone{figures/Figure6.eps}
\caption
{Effective $\alpha$'s for the $t_{cool} =$ 2 ORP, $\gamma = 5/3$, $\Sigma \sim r^{ -1}$ disk, averaged between 12 and 18 ORPs, for four different different azimuthal grid resolutions, given by the number of azimuthal grid-points Lmax. The parallel dashed lines show the predictions of Gammie (2001) for $\alpha$ (upper: strongly self-gravitating limit, lower: non-self-gravitating limit), while the dashed line at $\log(\alpha) = -1.22$ represents the maximum stress a disk can maintain without fragmenting (Rice et al. 2005) . 
}
\end{figure}
\newpage

\begin{figure}
\plotone{figures/Figure7.eps}
\caption
{Alpha variations.}
\end{figure}
